Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-17 manuscript.Rmd
very small tweaks.

We use data from 7 nationally representative panel surveys from Australia, Germany, Great Britain, Switzerland, and the United States (summarized in Table 1), combining all the data files used in previous work [@kiley2020; @lersch2023].[^1] These studies cover a long period of time (the range of surveys spans from 1968 to 2021), with 610 personal culture items that capture attitudes, beliefs, values, self-assessments, self-descriptions, and behaviors [@alwin2007]. We restricted the sample such that individuals between the ages of 18 and 79 are included without further elimination, and in all surveys, we used all possible cases for which respondents provided responses. In the end, the analyses that follow rely on a cross-cultural sample with a cross-domain set of items to capture the breadth of individual personal culture. Appendix A documents the list of all variables used in the upcoming analyses.

@@

table <- tibble(
  Country = c("Australia", "Germany", "Great Britian", "Great Britian", "Switzerland", "United States", "United States"),
  Survey = c("Household, Income and Labor Dynamics in Australia (HILDA)",
             "Socio-Economic Panel Study (SOEP)",
             "British Household Panel Survey (BHPS)",
             "Understanding Society/UK Household Longitudinal Study (UKHLS)",
             "Swiss Household Panel (SHP)",
             "General Social Survey (GSS)",
             "Panel Study of Income Dynamics (PSID)"),
  Period = c("2001-2021", "1984-2020", "1991-2008", "2009-2020", "1999-2019", "2006-2012", "1968-2019"),
  Outcomes = c(30, 122, 83, 66, 77, 183, 49)
)

@@

As a first step in our analysis, we fit LCAMs to each of the 610 measures of personal culture outlined above. We then use these models to derive two measures of residual variance that correspond to our theoretical processes. First, to measure the importance of interpersonal differences, we calculate the residuals derived from measuring each respondent at the midpoint of their age trajectory, denoted as $\tau$. By dividing this value by the total variance in responses and subtracting it from 1, we calculate the proportion of variance explained by interpersonal differences. Second, we calculate residuals for additionally accounting for the age trajectory, denoted as $\delta$. Subtracting the variance explained by interpersonal differences and dividing this new value by the total variance give us the additional proportion of variance explained by intrapersonal change. These two components reflect total “explained” variance of the model. Our principal outcome of interest is the proportion of explained variance accounted for by intrapersonal change, which we calculate as

@@

Figure 1 plots the proportion of variance explained by interpersonal differences (“between”) and intrapersonal change (“within”) for each of the 610 questions analyzed, plotted separately by panel. We highlight two top-line findings. First, despite the differences in interpretation present in previous works, results from the GSS appear remarkably consistent with results from the other panels. All panels show a wide range of the proportion of explained variance attributable to intrapersonal change, with all panels including questions that show essentially no variance explained by intrapersonal change and all panels including a few questions where more than 20 percent of explained variance is attributable to intrapersonal change. Across all the questions, interpersonal differences explain a much larger share of the variance in responses. Again, this is to be expected. Interpersonal differences capture not just pre-adult socialization, but all accumulated experiences up to the start of the panel.

The substantive importance of intrapersonal change depends on a range of factors, including how long the panel runs, whether assumptions about linear change hold, and whether the panel is capturing a distinctly turbulent period or a distinctly stable period for the item under question. However, we observe an average of about `PERCENT` percent of explained variance being attributed to intrapersonal change across all the questions, with an average duration of `YEARS`. If we assume that the period under observation is “typical” for a question – not a time of extremely heightened (or lowered) sensitivity or change – then it does not seem realistic for this process of intrapersonal change to explain a large share of interpersonal differences.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture, the attitudes, beliefs, values, and preferences that manifest at the individual level. Relying on an approach that quantified the amount of response variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions (such as factors that affect one’s life satisfaction or attitudes about government benefits), it is highly plausible that changes in personal culture as a result of peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small (`NUMBER?`) relative to interpersonal differences. On many questions, the amount of explained variance attributable to intrapersonal change is essentially zero. These include `EXAMPLES`. On these questions, assuming the window of observation is not uniquely stable, the amount of variance attributable to intrapersonal change does not seem sufficient to explain how and why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

@@

To be clear, this does not mean that adults do not change or are incapable of changing. On basically all questions, some members of the sample change their personal culture. And our results should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significnat political ramifications, even as most respondents remain fixed in their preferences.

@@

While we find that people continually reshape their personal culture over time, our results continue to direct attention to the “past conditions of production” that shape peoples’ personal culture early in life. To the extent that we wish to understand what makes adults different, focusing attention on peoples’ transition between social roles, selection into and out of organizational environments, and restructuring of social networks, as well as changes in the cultural resources people have at their disposal, seems to explain a much smaller proportion of their differences than the formative experiences that happen early in life.

The findings we present here are consonant with a range of recent causal inference work that suggests selection effects, rather than treatment effects, seem to explain the majority of differences in personal culture between people in different social roles or positions [@campbell2016; @wodtke2018]. While some people clearly change over time as they move into new roles, this amount of change does not appear to be sufficient to explain what are often substantial differences between people in different roles. Our results seem to suggest that, to the extent we observe differences in personal culture between people in different social roles, such as parenthood, having children, education, or working in positions of authority [@longest2013], or between people in different occupations [@weeden2005], our prior expectation should be that this difference is selection, rather than treatment.

@@

Because intrapersonal change in personal culture does not appear to be a driving force in cultural differences among adults, our results suggest that processes of social and cultural change at the aggregate level are likely driven by cohort replacement – one group of people with cultural dispositions existing the population and a different set of people with a potentially different set of cultural dispositions entering the population. This is likely true for both large-scale change in societies over time as well as changes within organizations, political parties, occupations, and other social aggregates.

@@

A major takeaway from the meta-analysis is that greater response resolution – the number of response options survey respondents are given – is associated with a greater share of variance being explained by within-individual change. Again, it is possible that this reflects different kinds of issues being measured with different kinds of scales. But it could also suggest that within-person change, when it happens, is relatively small – moving from “agree” to “strongly agree”, rather than “strongly agree” to “strongly disagree,” for example – and therefore easier to detect with greater resolution.

Second, our results continue to highlight the value of panel surveys for adjudicating different theories of culture that cannot be evaluated with cross-sectional data alone. The process of how individual people store and report personal culture is key to understand the aggregate processes of social change, and . Additionally, we echo previous calls for studies – including panel studies – of the socialization processes of young people. Our results suggest that much of the work of social differentiation takes place before people turn 18. In order to understand why adults are different, we need to understand them before they are adults.

>>>>

We use data from 7 nationally representative panel surveys from Australia, Germany, Great Britain, Switzerland, and the United States (summarized in Table 1), combining all the data files used in previous work [@kiley2020; @lersch2023].[^1] These studies cover a long period of time (the range of surveys spans from 1968 to 2021), with 609 personal culture items that capture attitudes, beliefs, values, self-assessments, self-descriptions, and behaviors [@alwin2007]. We restricted the sample such that individuals between the ages of 18 and 79 are included without further elimination, and in all surveys, we used all possible cases for which respondents provided responses. In the end, the analyses that follow rely on a cross-cultural sample with a cross-domain set of items to capture the breadth of individual personal culture. Appendix A documents the list of all variables used in the upcoming analyses.

@@

table <- tibble(
  Country = c("Australia", "Germany", "Great Britian", "Great Britian", "Switzerland", "United States", "United States"),
  Survey = c("Household, Income and Labor Dynamics in Australia (HILDA)",
             "Socio-Economic Panel Study (SOEP)",
             "British Household Panel Survey (BHPS)",
             "Understanding Society/UK Household Longitudinal Study (UKHLS)",
             "Swiss Household Panel (SHP)",
             "General Social Survey (GSS)",
             "Panel Study of Income Dynamics (PSID)"),
  Period = c("2001-2021", "1984-2020", "1991-2008", "2009-2020", "1999-2019", "2006-2012", "1968-2019"),
  Outcomes = c(30, 122, 83, 66, 77, 183, 48)
)

@@

As a first step in our analysis, we fit LCAMs to each of the 609 measures of personal culture outlined above. We then use these models to derive two measures of residual variance that correspond to our theoretical processes. First, to measure the importance of interpersonal differences, we calculate the residuals derived from measuring each respondent at the midpoint of their age trajectory, denoted as $\tau$. By dividing this value by the total variance in responses and subtracting it from 1, we calculate the proportion of variance explained by interpersonal differences. Second, we calculate residuals for additionally accounting for the age trajectory, denoted as $\delta$. Subtracting the variance explained by interpersonal differences and dividing this new value by the total variance give us the additional proportion of variance explained by intrapersonal change. These two components reflect total “explained” variance of the model. Our principal outcome of interest is the proportion of explained variance accounted for by intrapersonal change, which we calculate as

@@

Figure 1 plots the proportion of variance explained by interpersonal differences (“between”) and intrapersonal change (“within”) for each of the 609 questions analyzed, plotted separately by panel. We highlight two top-line findings. First, despite the differences in interpretation present in previous works, results from the GSS appear remarkably consistent with results from the other panels. All panels show a wide range of the proportion of explained variance attributable to intrapersonal change, with all panels including questions that show essentially no variance explained by intrapersonal change and all panels including a few questions where more than 20 percent of explained variance is attributable to intrapersonal change. Across all the questions, interpersonal differences explain a much larger share of the variance in responses. Again, this is to be expected. Interpersonal differences capture not just pre-adult socialization, but all accumulated experiences up to the start of the panel.

The substantive importance of intrapersonal change depends on a range of factors, including how long the panel runs, whether assumptions about linear change hold, and whether the panel is capturing a distinctly turbulent period or a distinctly stable period for the item under question. However, we observe an average of about 9.7 percent of explained variance being attributed to intrapersonal change across all the questions. If we assume that the period under observation is “typical” for a question –-not a time of extremely heightened (or lowered) sensitivity or change-– then it does not seem realistic for this process of intrapersonal change to explain a large share of interpersonal differences.

@@

This paper set out to clarify previous findings about the prevalence of durable change in personal culture -- the attitudes, beliefs, values, and preferences that manifest at the individual level. Relying on an approach that quantified the amount of response variance attributable to either intrapersonal change or interpersonal differences at baseline, we found a consistent pattern across data sets previously used in this debate. We see evidence of intrapersonal change over time, with some measures of personal culture displaying quite high levels of intrapersonal change relative to the amount of variance between people. On some questions (such as factors that affect one’s life satisfaction or attitudes about government benefits), it is highly plausible that changes in personal culture as a result of peoples’ experiences as adults are the primary explanation for differences in personal culture.

However, the amount of intrapersonal change is often small (approximately 10% across all surveys) relative to interpersonal differences. On many questions, the amount of explained variance attributable to intrapersonal change is essentially zero. These include questions about civil liberties, abortion, interindividual trust, or sense of civic duty. On these questions, assuming the window of observation is not uniquely stable, the amount of variance attributable to intrapersonal change does not seem sufficient to explain how and why adults end up with different personal cultures. For these items, the explanation for personal cultural differences seems to reside principally in early-life experiences.

@@

To be clear, this does not mean that adults do not change or are incapable of changing. On basically all questions, some members of the sample change their personal culture. And our results should not be taken to imply that intrapersonal change is not important or consequential. Small numbers of changers can have important consequences in the aggregate. A shift from 49 percent supporting gay marriage to 51 percent supporting it can have significant political ramifications, even as most respondents remain fixed in their preferences.

@@

While we find that people continually reshape their personal culture over time, our results continue to direct attention to the “past conditions of production” that shape peoples’ personal culture early in life. To the extent that we wish to understand what makes adults different, focusing attention on peoples’ transition between social roles, selection into and out of organizational environments, and restructuring of social networks, as well as changes in the cultural resources people have at their disposal, seem to explain a much smaller proportion of their differences than the formative experiences that happen early in life.

The findings we present here are consonant with a range of recent causal inference work that suggests selection effects, rather than treatment effects, seem to explain the majority of differences in personal culture between people in different social roles or positions [@campbell2016; @wodtke2018]. While some people clearly change over time as they move into new roles, this amount of change does not appear to be sufficient to explain what are often substantial differences between people in different roles. Our results suggest that, to the extent we observe differences in personal culture between people in different social roles, such as parenthood, having children, education, or working in positions of authority [@longest2013], or between people in different occupations [@weeden2005], our prior expectation should be that this difference is selection, rather than treatment.

@@

Because intrapersonal change in personal culture does not appear to be a driving force in cultural differences among adults, our results suggest that processes of social and cultural change at the aggregate level are likely driven by cohort replacement –- one group of people with certain cultural dispositions exiting the population and a different set of people with a potentially different set of cultural dispositions entering the population [@vaisey2016]. This is likely true for both large-scale change in societies over time as well as changes within organizations, political parties, occupations, and other social aggregates.

@@

A major takeaway from the meta-analysis is that greater response resolution –-the number of response options survey respondents are given-– is associated with a greater share of variance being explained by within-individual change. Again, it is possible that this reflects different kinds of issues being measured with different kinds of scales. But it could also suggest that within-person change, when it happens, is relatively small –-moving from “agree” to “strongly agree”, rather than “strongly agree” to “strongly disagree,” for example-– and therefore easier to detect with greater resolution.

Second, our results continue to highlight the value of panel surveys for adjudicating different theories of culture that cannot be evaluated with cross-sectional data alone. The process of how individual people store and report personal culture is key to understand the aggregate processes of social change.

Additionally, we echo previous calls for studies –-including panel studies-– of the socialization processes of young people. Our results suggest that much of the work of social differentiation takes place before people turn 18. In order to understand why adults are different, we need to understand them before they are adults.
